---
title: "modeling graduation rates at US colleges using data from the {ISLR2} package"
author: "Keith Karani"
format: html
editor: visual
---

## Guiding questions

1.  Do colleges with large full-time enrollments have lower graduation rates?

2.  How do graduation rate compare between public and private colleges?

3.  Is the graduation rate different for selective schools?

4.  Are there any correlations between graduation rates and others variables such as students-faculty ratio, tuition or student demographics

::: callout-important
### Source

This dataset was taken from the StatLib library which is maintained at Carnegie Mellon University. The dataset was used in the ASA Statistical Graphics Section's 1995 Data Analysis Exposition

### References

James, G., Witten, D., Hastie, T., and Tibshirani, R. (2013) *An Introduction to Statistical Learning with applications in R*, [https://www.statlearning.com](https://www.statlearning.com/), Springer-Verlag, New York
:::

```{r}
# setup chunk
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(ISLR2)

```

Lets learn about the data:

U.S. News and World Report's College Data: statistics for a large number of US Colleges from the year 1995 issue of US News and World Report

```{r}

# to learn more about the data from the its data dictionary
?College

glimpse(College)


summary(College)
```

Lets begin with some simple exploratory visualization to view our data in graphic representation.

we will begin with plotting a histogram showing the distribution of the graduation rate

```{r}

ggplot(College, aes(x = Grad.Rate)) +
  geom_histogram() +
  labs(
    x = "Graduation Rate"
  )
  theme_minimal()


#Observation: there is an outlier that is over 100%
  
# let's investigate the observed outlier
  
 gr_outlier <- filter(College, Grad.Rate >= 100 )
 
 gr_outlier
 
 # with the two primary variables that we will focus on lets create a scatter plot and observe

 ggplot(College, aes(x = log10(F.Undergrad), y = Grad.Rate)) +
   geom_point() +
   labs(
     caption = "Source: StatLib library which is maintained at Carnegie Mellon University",
     x = "Number of fulltime undergraduates",
     y = "Graduation Rate"
   ) +
   theme_classic() 
 
```

```{r}

college_small <- College |> 
  mutate(
    log_full = log10(F.Undergrad)
  ) |> 
  select(Grad.Rate, log_full, Private, Top25perc)
  
View(college_small)

```

Lets begin some modelling

```{r}

 ggplot(College, aes(x = log10(F.Undergrad), y = Grad.Rate)) +
   geom_point() +
   labs(
     caption = "Source: StatLib library which is maintained at Carnegie Mellon University",
     x = "Number of fulltime undergraduates",
     y = "Graduation Rate"
   ) +
   theme_classic() +
   geom_smooth(method = "lm")

 
# we can observe that the variable Full Time undergraduate population is not a good predictor for the Graduation Rate answering our first question
 
 
model_undergrad <- lm(Grad.Rate ~ log_full, data = college_small) 

summary(model_undergrad)

#diagnotic plot
plot(model_undergrad)




```

What about Private(A factor with levels `No` and `Yes` indicating private or public university)

```{r}

 ggplot(College, aes(x = log10(F.Undergrad), y = Grad.Rate, color = Private)) +
   geom_point() +
   labs(
     caption = "Source: StatLib library which is maintained at Carnegie Mellon University",
     x = "Number of fulltime undergraduates",
     y = "Graduation Rate"
   ) +
   theme_classic() +
   geom_smooth(method = "lm", se = FALSE) +
   scale_color_brewer(palette = "Dark2")

# lets model this
 
 model_private <- lm(Grad.Rate ~ Private + log_full, data = college_small)
 
 summary(model_private)
 
 
# Interaction: privare and Full time undergrad
 
model_private_int <- lm(Grad.Rate ~ Private * log_full, data = college_small)

summary(model_private_int)

# Compute analysis of variance (or deviance) tables for one or more fitted model objects.
anova(model_private_int)

```

```{r}
# what about Top 25%

model_top <- lm(Grad.Rate ~ Private + log_full + Top25perc, data = college_small)



```
